---
title: "Human CRPS Combined Cohorts 16S Analysis"
author: "Rachel Rodgers & Marie Crane"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=8,
                      fig.height=6,
                      warning = FALSE,
                      message = FALSE)
```

```{r load-libraries, message=FALSE, include=FALSE}
library("phyloseq")
library("vegan")
library("microbiome")
library("picante")
library("ggrepel")
library("ggpubr")
library("gridExtra")
library("scales")
library("tidyverse")

source("./shared_R_scripts/Helper_Functions.R")
```

# Overview

---

Analysis of combined human CRPS cohorts:
  * Run 1: 190201 (6 samples)
  * Run 2: 190318 (10 samples)
  * Run 3: 190523 (10 samples)
  * Run 4: 200127 (20 samples)
  * Run 5: 201019 (24 samples)

These cohorts include acute and chronic CRPS patients with their matched household controls, as well as unrelated, healthy "BioBank" controls.

Analyzed:

1. Three groups ("sample_type" groups):
  * CRPS
  * HHC
  * BioBank ("healthy")


2. Six groups ("crps_match" groups):
  * Acute CRPS
  * Chronic CRPS
  * Acute HHC
  * Chronic HHC
  * Acute-matched BioBank
  * Chronic-matched BioBank

----

```{r read-in-data}
physeqRaw <- readRDS("../data/physeqObjects/ps0.rdp_single.RDS")
physeqRaw # 2115 taxa x 69 samples

sampleData <- readRDS("../data/sampleDataModified.RDS")
```

```{r physeqMerged}
# add readsPerSample to sample data prior to merging
readsPerSample <- sample_sums(physeqRaw)
sampleData <- mutate(sampleData, 
                     "readsPerSample" = readsPerSample[sample])

row.names(sampleData) <- sampleData$sample
physeqMerged <- merge_phyloseq(physeqRaw, sample_data(sampleData))
physeqMerged # 2115 x 69
```

# Quality Control

## Sample Outlier Removal by Unexpected Read Count

Checking the average read count across all samples to determine if any have unexpected numbers of reads.

```{r fxn-PlotAvgReads}
# sample_type = 3 groups
# crps_match = 6 groups
groupCategories <- c("sample type" = "sample_type",
                     "CRPS match" = "crps_match")

PlotAvgReads <- function(df, xVar, yVar, groupVar, 
                         plotTitle = NULL, thresh = 1000,
                         labelSamples = FALSE,
                         includeStats = FALSE) {
  
  basePlot <- ggplot(df, aes_string(x = xVar, y = yVar, fill = groupVar)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.75) +
    geom_hline(yintercept = thresh, lty = 2) +
    ylab("Reads per Sample") +
    ggtitle(plotTitle) +
    scale_y_log10() +
    theme_pubr() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5),
          axis.title.x = element_blank(),
          axis.text.x = element_text(size = 10)) 
  
  if (labelSamples == TRUE) {
    basePlot <- basePlot + 
      geom_text_repel(aes(label = paste(sample, ": ", readsPerSample)),
                    data = subset(df, readsPerSample <= thresh))
  }
  
  if (includeStats == TRUE) {
    basePlot <- basePlot +
      stat_compare_means(method = "kruskal.test", label.x.npc = 0)
  }
  
  return(basePlot)

}
```

```{r avgReadsPlot, fig.width=11, fig.height=8.5}
avgReadsPlots <- pmap(.l = list(xVar = groupCategories,
                                groupVar = groupCategories,
                                plotTitle = c("sample type" = "Average Reads by Sample Type",
                                              "CRPS match" = "Average Reads by CRPS Matched Groups")),
                      .f = PlotAvgReads,
                      df = sampleData, yVar = "readsPerSample",
                      labelSamples = TRUE, includeStats = FALSE)

ggarrange(plotlist = avgReadsPlots, ncol = 1, nrow = 2)
```

From the box plots of average reads per group, we can see there are several BioBank control samples and a technical control with low reads.  These samples will be dropped from analysis.

```{r physeqTrimmed}
lowReadSamples <- sampleData %>% 
  filter(readsPerSample <= 1000) %>% 
  pull(sample) %>% 
  as.character()

physeqTrimmed <- physeqMerged %>% 
  subset_samples(!(sample %in% lowReadSamples)) %>% 
  RemoveMissingTaxa()
physeqTrimmed # 2108 x 64 samples
```

Removal of samples having fewer than 1000 reads reduced the total samples from 69 to 64, and total taxa from 2,115 to 2,108.

Replotting average reads per group to ensure there is no significant difference in average read numbers:

```{r avgReadsPlots-after-filtering, fig.width=11, fig.height=8.5}
sampleDataTrimmed <- as(sample_data(physeqTrimmed), "data.frame")

avgReadsPlotsTrimmed <- pmap(.l = list(xVar = groupCategories,
                                       groupVar = groupCategories,
                             plotTitle = c("sample type" = "Average Reads by Sample Type",
                                           "CRPS match" = "Average Reads by CRPS Matched Groups")),
                             .f = PlotAvgReads,
                             df = sampleDataTrimmed, yVar = "readsPerSample",
                             labelSamples = FALSE, includeStats = TRUE)

ggarrange(plotlist = avgReadsPlotsTrimmed, ncol = 1, nrow = 2)
```

Because a significant difference still exists between groups, we will rarefy the data set downstream.

## Taxa Filtering

Non-bacterial taxa need to be removed from the data set.

```{r filter-non-bacterial}
physeqBacteria <- physeqTrimmed %>% 
  subset_taxa(Kingdom == "Bacteria" & Phylum != "Cyanobacteria/Chloroplast")

# did this deplete any samples?
any(sample_sums(physeqBacteria) == 0) # no

physeqBacteria
```

Removal of non-bacterial taxa furhter reduces total taxa in the data set from 2,108 to 2,054.

## Rarefy

```{r rarefy}
physeqBacteriaRare <- rarefy_even_depth(physeqBacteria, rngseed = 2236001)
physeqBacteriaRare # 1,612 taxa
```

```{r save-rarefy}
saveRDS(physeqBacteriaRare, file = "../data/physeqObjects/physeqBacteriaRare.RDS")
```

