---
title: "Human CRPS Combined Cohorts 16S Analysis (Rarefied)"
author: "Rachel Rodgers"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=8,
                      fig.height=6,
                      warning = FALSE,
                      message = FALSE)
```

```{r load-libraries, message=FALSE, include=FALSE}
library("shiny")
library("phyloseq")
library("vegan")
library("gridExtra")
library("knitr")
library("plotly")
library("microbiome")
library("data.table")
library("ggrepel")
library("ggpubr")
library("scales")
library("plyr")
library("DT")
library("picante")
library("gridExtra")
library("tidyverse")

source("./shared_R_scripts/Helper_Functions.R")

options(shiny.sanitize.errors = FALSE)
```

## Overview {.tabset}

---

Analysis of combined human CRPS cohorts:

  * Run 1: 190201 (6 samples)
  * Run 2: 190318 (10 samples)
  * Run 3: 190523 (10 samples)
  * Run 4: 200127 (20 samples)
  * Run 5: 201019 (24 samples)

These cohorts include acute and chronic CRPS patients with their matched household controls, as well as unrelated, healthy "BioBank" controls.

Analyzed:

1. Three groups ("sample_type" groups):
  * CRPS
  * HHC
  * BioBank ("healthy")


2. Six groups ("crps_match" groups):
  * Acute CRPS
  * Chronic CRPS
  * Acute HHC
  * Chronic HHC
  * Acute-matched BioBank
  * Chronic-matched BioBank
  
3. Comparisons to BioBank (Non-related) Controls
  * all CRPS vs all BioBank 
  * Chronic CRPS vs Chronic-matched BioBank
  * Acute CRPS vs Chronic-matched BioBank

----

```{r read-in-data, include=FALSE}
physeqRaw <- readRDS("../data/physeqObjects/ps0.rdp_single.RDS")
physeqRaw # 2115 taxa x 69 samples

sampleData <- readRDS("../data/sampleDataModified.RDS")
```

```{r physeqMerged, results='hide', include=FALSE}
# add readsPerSample to sample data prior to merging
readsPerSample <- sample_sums(physeqRaw)
sampleData <- mutate(sampleData, 
                     "readsPerSample" = readsPerSample[sample])

row.names(sampleData) <- sampleData$sample
physeqMerged <- merge_phyloseq(physeqRaw, sample_data(sampleData))
physeqMerged # 2115 x 69
```

### **Quality Control**

#### **Sample Outlier Removal by Unexpected Read Count**

Checking the average read count across all samples to determine if any have unexpected numbers of reads.

```{r fxn-PlotAvgReads}
# sample_type = 3 groups
# crps_match = 6 groups
groupCategories <- c("sample type" = "sample_type",
                     "CRPS match" = "crps_match")

PlotAvgReads <- function(df, xVar, yVar, groupVar, 
                         plotTitle = NULL, thresh = 1000,
                         labelSamples = FALSE,
                         includeStats = FALSE) {
  
  basePlot <- ggplot(df, aes_string(x = xVar, y = yVar, fill = groupVar)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.75) +
    geom_hline(yintercept = thresh, lty = 2) +
    ylab("Reads per Sample") +
    ggtitle(plotTitle) +
    scale_y_log10() +
    theme_pubr() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5),
          axis.title.x = element_blank(),
          axis.text.x = element_text(size = 10)) 
  
  if (labelSamples == TRUE) {
    basePlot <- basePlot + 
      geom_text_repel(aes(label = paste(sample, ": ", readsPerSample)),
                    data = subset(df, readsPerSample <= thresh))
  }
  
  if (includeStats == TRUE) {
    basePlot <- basePlot +
      stat_compare_means(method = "kruskal.test", label.x.npc = 0)
  }
  
  return(basePlot)

}
```

```{r avgReadsPlot, fig.width=11, fig.height=8.5}
avgReadsPlots <- pmap(.l = list(xVar = groupCategories,
                                groupVar = groupCategories,
                                plotTitle = c("sample type" = "Average Reads by Sample Type",
                                              "CRPS match" = "Average Reads by CRPS Matched Groups")),
                      .f = PlotAvgReads,
                      df = sampleData, yVar = "readsPerSample",
                      labelSamples = TRUE, includeStats = FALSE)

ggarrange(plotlist = avgReadsPlots, ncol = 1, nrow = 2)
```

From the box plots of average reads per group, we can see there are several BioBank control samples and a technical control with low reads.  These samples will be dropped from analysis.

```{r physeqTrimmed}
lowReadSamples <- sampleData %>% 
  filter(readsPerSample <= 1000) %>% 
  pull(sample) %>% 
  as.character()

keepSamples <- sampleData$sample[!(sampleData$sample %in% lowReadSamples)]

physeqTrimmed <- prune_samples(keepSamples, physeqMerged) %>% 
  RemoveMissingTaxa()

physeqTrimmed # 2108 x 64 samples
```

Removal of samples having fewer than 1000 reads reduced the total samples from 69 to 64, and total taxa from 2,115 to 2,108.

Replotting average reads per group to ensure there is no significant difference in average read numbers:

```{r avgReadsPlots-after-filtering, fig.width=11, fig.height=8.5}
sampleDataTrimmed <- as(sample_data(physeqTrimmed), "data.frame")

avgReadsPlotsTrimmed <- pmap(.l = list(xVar = groupCategories,
                                       groupVar = groupCategories,
                             plotTitle = c("sample type" = "Average Reads by Sample Type",
                                           "CRPS match" = "Average Reads by CRPS Matched Groups")),
                             .f = PlotAvgReads,
                             df = sampleDataTrimmed, yVar = "readsPerSample",
                             labelSamples = FALSE, includeStats = TRUE)

ggarrange(plotlist = avgReadsPlotsTrimmed, ncol = 1, nrow = 2)
```

Because a significant difference still exists between groups, we will rarefy the data set.

```{r rarefy, results='hide'}
# Rarefy
physeqRare <- rarefy_even_depth(physeqTrimmed, rngseed = 2236001)
physeqRare # 1,612 taxa
```

```{r save-rarefy, eval=FALSE}
saveRDS(physeqRare, file = "../data/physeqObjects/physeqRare.RDS")
```

#### **Taxa Filtering & Abundance**

Non-bacterial taxa need to be removed from the data set.

```{r filter-non-bacterial, results='hide'}
physeqBacteria <- physeqRare %>% 
  subset_taxa(Kingdom == "Bacteria" & Phylum != "Cyanobacteria/Chloroplast")

# did this deplete any samples?
any(sample_sums(physeqBacteria) == 0) # no

physeqBacteria
```

Removal of non-bacterial taxa further reduces total taxa in the data set from 2,108 to 2,054.

Checking the prevalence and abundance values of different bacterial families to see if there are any low abundance/prevalence taxa.

```{r phylaPrevalence, fig.width=9, fig.height=6}
phylaPrevalence <- TaxRankPrevalence(physeqBacteria, "Phylum")
phylaPrevalence$prevalence_table

phylaPrevalencePlot <- ggplot(phylaPrevalence$prevalence_df,
                              aes(x = TotalAbundance,
                                  y = Prevalence/nsamples(physeqBacteria),
                                  color = Family)) +
  geom_hline(yintercept = 1/207, alpha = 0.5, linetype = 2) +
  geom_point(size = 3, alpha = 0.7) +
  scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~ Phylum) +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) +
  ggtitle("Phylum Prevalence in All Samples",
          subtitle = "Colored by Familiy")
phylaPrevalencePlot
```

From the prevalence plots, we can see that several low-abundance/low-prevalence taxa are present in the datset, including the Fusocbacteria, Lentisphaerae, and Syngeristetes.  Downstream analysis (such as biomarker analysis) may benefit from removal of these taxa.

```{r, save-physeqBacteria-for-lefse-input, eval=FALSE, include=FALSE}
# Save the physeqBacteria oject to generate LEfSe input data
saveRDS(physeqBacteria, "../data/physeqObjects/physeqBacteria.RDS")
```

```{r analysis-globals}
# Data that can be used for both 3 and 6 group comparisons

#----- Alpha Diversity DF -----#

sampleDataBacteria <- as(sample_data(physeqBacteria), "data.frame")

alphaIdxList <- list("Richness" = "observed",
                     "Shannon Diversity" = "diversity_shannon", 
                     "Faith's PD" = "PD")

alphaDiv <- microbiome::alpha(physeqBacteria, index = c("observed",
                                                        "diversity_shannon"))

faithsPD <- pd(as.data.frame(physeqBacteria@otu_table),
               physeqBacteria@phy_tree,
               include.root = FALSE)

allAlphaDiv <- merge(alphaDiv, faithsPD, by = "row.names", all = TRUE)
allAlphaDiv <- column_to_rownames(allAlphaDiv, var = "Row.names")

sampleDataBacteria <- merge(allAlphaDiv, sampleDataBacteria,
                            by = "row.names", all = TRUE) %>% 
  column_to_rownames(var = "Row.names")
```

### **16S Analysis (3 groups)**

Three groups:

  * CRPS
  * HHC
  * BioBank Control

#### **Community Composition Plot**

```{r comm-comp-3groups}
#----- UI -----#

inputPanel(
  # select taxonomic rank
  selectInput(inputId = "commTaxRank3", label = "Taxonomic Rank",
              choices = c("Kingdom", "Phylum", "Class", "Order", "Family",
                          "Genus", "Species"),
              selected = "Phylum"),
  # select abundance filter
  sliderInput(inputId = "commAbdFilter3", label = "Abundance Filter",
              min = 0, max = 1, step = 0.05, value = 0),
  # hide NA?
  checkboxInput(inputId = "commHideNA3", label = "Hide unnamed taxa", 
                value = FALSE)
)

# update plot:
actionButton(inputId = "commUpdate3", label = "Update Composition Plot",
             icon = icon("fas fa-sync-alt"))

#----- React -----#

CalculateAbundance3 <- eventReactive(input$commUpdate3, {
  
  withProgress(message = "Calculating Abundance...", value = 1, {
                 abdDF <- MakeAbundanceDF(physeq = physeqBacteria,
                                          taxRank = input$commTaxRank3,
                                          abundanceFilter = input$commAbdFilter3,
                                          pruneMissing = input$commHideNA3)

                 if (dim(abdDF)[1] == 0) {
                   stop("No taxa meet this filtering criteria. 
                        Try lowering the Abundance Filter option.",
                        call. = FALSE)
                 } else {
                   if (input$commTaxRank3 == "Species") {
                     
                     abdDF <- mutate(abdDF,
                                     "GenusSpecies" = paste(Genus, Species, 
                                                            sep = " "),
                                     no = "other")
                   }
                 }
                 abdDF
               })
}, ignoreNULL = FALSE)


renderPlotly({
  compPlot <- PlotCommunityComposition(abdDF = CalculateAbundance3(),
                                       taxRank = ifelse(isolate(input$commTaxRank3) == "Species",
                                                        yes = "GenusSpecies",
                                                        no = isolate(input$commTaxRank3)),
                                       facetFormula = "~ sample_type")
  
  ggplotly(compPlot, height = 600, width = 1000)
})

```

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>


#### **Alpha Diversity**

```{r alpha-div-3groups, fig.width=11, fig.height=11}
alphaDivPlots3 <- pmap(.l = list(yVar = alphaIdxList,
                                 yLab = names(alphaIdxList)),
                       .f = PlotAlphaDiversity,
                       df = sampleDataBacteria,
                       xVar = "sample_type",
                       statMethod = "kruskal.test")

grid.arrange(alphaDivPlots3$Richness,
             alphaDivPlots3$`Shannon Diversity`,
             alphaDivPlots3$`Faith's PD`,
             ncol = 2, nrow = 2)
```

<br>
<br>
<br>

#### **Beta Diversity**

```{r beta-div-3groups}
#----- UI -----#

inputPanel(
  # select distance method
  radioButtons(inputId = "betaDist3", label = "Distance Method:",
               choices = c("UniFrac" = "unifrac",
                           "Weighted UniFrac" = "wunifrac"),
               selected = "unifrac"),
  # show sample names
  checkboxInput(inputId = "betaShowNames3", label = "Show point labels",
                value = FALSE),
  # show confidence ellipses
  checkboxInput(inputId = "betaShowEllipses3", 
                label = "Show confidence ellipses", value = TRUE)
)

#----- React -----#

CalculateOrdination3 <- reactive({
  withProgress(message = "Calculating ordination (3 groups)...", value = 1, {
    # Generate ordination object
    set.seed(1648370)
    ordinationObj <- ordinate(physeqBacteria, method = "PCoA",
                              distance = input$betaDist3)
  })
})

renderPlot({
  basePlot <- plot_ordination(physeqBacteria,
                              ordination = CalculateOrdination3(),
                              color = "sample_type", shape = "sample_type") +
    theme_bw() +
    theme(legend.title = element_text(size = 18),
          legend.text = element_text(size = 16),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          axis.text.x = element_text(size = 12),
          axis.text.y = element_text(size = 12)) +
    geom_point(size = 4)
  
  if (input$betaShowNames3) {
    basePlot <- basePlot +
      geom_text_repel(aes_string(label = "sample"), color = "grey30", size = 4)
  }
  
  if (input$betaShowEllipses3 == TRUE) {
    basePlot <- basePlot + stat_ellipse(type = "norm")
  }
  
  basePlot
  
}, height = 600, width = 900)

```

**ADONIS Results for selected Week and Distance Measure:**

```{r beta-adonis-3groups}
set.seed(1648370)
renderPrint(
  RunAdonis(physeqObj = physeqBacteria,
            category = "sample_type", distance = input$betaDist3)
)
```

### **16S Analysis (6 groups)**

Six groups ("crps_match" groups):

  * Acute CRPS (CRPSA)
  * Chronic CRPS (CRPSC)
  * Acute HHC (HHCA)
  * Chronic HHC (HHCC)
  * Acute-matched BioBank (NRAM)
  * Chronic-matched BioBank (NRCM)
  
#### **Community Composition Plot**

```{r comm-comp-6groups}
#----- UI -----#

inputPanel(
  # select taxonomic rank
  selectInput(inputId = "commTaxRank6", label = "Taxonomic Rank",
              choices = c("Kingdom", "Phylum", "Class", "Order", "Family",
                          "Genus", "Species"),
              selected = "Phylum"),
  # select abundance filter
  sliderInput(inputId = "commAbdFilter6", label = "Abundance Filter",
              min = 0, max = 1, step = 0.05, value = 0),
  # hide NA?
  checkboxInput(inputId = "commHideNA6", label = "Hide unnamed taxa", 
                value = FALSE)
)

# update plot:
actionButton(inputId = "commUpdate6", label = "Update Composition Plot",
             icon = icon("fas fa-sync-alt"))

#----- React -----#

CalculateAbundance6 <- eventReactive(input$commUpdate6, {
  
  withProgress(message = "Calculating Abundance...", value = 1, {
                 abdDF <- MakeAbundanceDF(physeq = physeqBacteria,
                                          taxRank = input$commTaxRank6,
                                          abundanceFilter = input$commAbdFilter6,
                                          pruneMissing = input$commHideNA6)

                 if (dim(abdDF)[1] == 0) {
                   stop("No taxa meet this filtering criteria. 
                        Try lowering the Abundance Filter option.",
                        call. = FALSE)
                 } else {
                   if (input$commTaxRank6 == "Species") {
                     
                     abdDF <- mutate(abdDF,
                                     "GenusSpecies" = paste(Genus, Species, 
                                                            sep = " "),
                                     no = "other")
                   }
                 }
                 abdDF
               })
}, ignoreNULL = FALSE)


renderPlotly({
  compPlot <- PlotCommunityComposition(abdDF = CalculateAbundance6(),
                                       taxRank = ifelse(isolate(input$commTaxRank6) == "Species",
                                                        yes = "GenusSpecies",
                                                        no = isolate(input$commTaxRank6)),
                                       facetFormula = "~ crps_match")
  
  ggplotly(compPlot, height = 600, width = 1000)
})

```

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>


#### **Alpha Diversity**

```{r alpha-div-6groups, fig.width=11, fig.height=11}
alphaDivPlots6 <- pmap(.l = list(yVar = alphaIdxList,
                                 yLab = names(alphaIdxList)),
                       .f = PlotAlphaDiversity,
                       df = sampleDataBacteria,
                       xVar = "crps_match",
                       statMethod = "kruskal.test")

grid.arrange(alphaDivPlots6$Richness,
             alphaDivPlots6$`Shannon Diversity`,
             alphaDivPlots6$`Faith's PD`,
             ncol = 2, nrow = 2)
```

<br>
<br>
<br>

#### **Beta Diversity**

```{r beta-div-6groups}
#----- UI -----#

inputPanel(
  # select distance method
  radioButtons(inputId = "betaDist6", label = "Distance Method:",
               choices = c("UniFrac" = "unifrac",
                           "Weighted UniFrac" = "wunifrac"),
               selected = "unifrac"),
  # show sample names
  checkboxInput(inputId = "betaShowNames6", label = "Show point labels",
                value = FALSE),
  # show confidence ellipses
  checkboxInput(inputId = "betaShowEllipses6", 
                label = "Show confidence ellipses", value = TRUE)
)

#----- React -----#

CalculateOrdination6 <- reactive({
  withProgress(message = "Calculating ordination (6 groups)...", value = 1, {
    # Generate ordination object
    set.seed(1648370)
    ordinationObj <- ordinate(physeqBacteria, method = "PCoA",
                              distance = input$betaDist6)
  })
})

renderPlot({
  basePlot <- plot_ordination(physeqBacteria,
                              ordination = CalculateOrdination6(),
                              color = "sample_type", shape = "sample_type") +
    theme_bw() +
    theme(legend.title = element_text(size = 18),
          legend.text = element_text(size = 16),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          axis.text.x = element_text(size = 12),
          axis.text.y = element_text(size = 12)) +
    geom_point(size = 4)
  
  if (input$betaShowNames6) {
    basePlot <- basePlot +
      geom_text_repel(aes_string(label = "sample"), color = "grey30", size = 4)
  }
  
  if (input$betaShowEllipses6 == TRUE) {
    basePlot <- basePlot + stat_ellipse(type = "norm")
  }
  
  basePlot
  
}, height = 600, width = 900)

```

**ADONIS Results for selected Week and Distance Measure:**

```{r beta-adonis-6groups}
set.seed(1648370)
renderPrint(
  RunAdonis(physeqObj = physeqBacteria,
            category = "crps_match", distance = input$betaDist6)
)
```

<br>

### **BioBank Comparisons**

Comparisons to BioBank (Non-related) Controls:

  * all CRPS vs all BioBank 
  * Chronic CRPS vs Chronic-matched BioBank
  * Acute CRPS vs Chronic-matched BioBank
  
```{r biobank-globals}

#----- physeqSubsetList -----#

# Subsets:
#   All CRPS & All BioBank (physeqCRPSBB)
#   Chronic CRPS & Chronic BioBank (physeqChronicBB)
#   Acute CRPS & Acute BioBank (physeqAcuteBB)

comparisonLabels <- c("CRPS_BB", "Chronic_BB", "Acute_BB")

conditionVecList <- list(c("CRPS", "BioBank_Control"),
                         c("CRPSC", "NRCM"),
                         c("CRPSA", "NRAM"))
names(conditionVecList) <- comparisonLabels

columnVarList <- list("sample_type",
                      "crps_match",
                      "crps_match")
names(columnVarList) <- comparisonLabels

physeqSubsetList <- vector(mode = "list", length = length(comparisonLabels))

samplesByConditionList <- vector(mode = "list", length = length(comparisonLabels))

for (i in 1:length(comparisonLabels)) {
  
  comparison <- comparisonLabels[i]
  conditionVec <- conditionVecList[[comparison]]
  column <- columnVarList[[comparison]]
  
  keepSamples <- sampleDataBacteria %>% 
    filter(!!(sym(column)) %in% conditionVec) %>% 
    pull(sample)
  
  samplesByConditionList[[i]] <- keepSamples
  names(samplesByConditionList)[i] <- comparison
  
  psSubset <- prune_samples(keepSamples, physeqBacteria) %>% 
    RemoveMissingTaxa()
  
  physeqSubsetList[[i]] <- psSubset
  names(physeqSubsetList)[i] <- comparison
}

#----- Alpha Diversity DFs -----#

alphaDivDFList <- vector(mode = "list", length = length(comparisonLabels))

for (i in 1:length(comparisonLabels)) {
  
  comparison <- comparisonLabels[i]
  samples <- samplesByConditionList[[comparison]]
  
  sampleDataFilt <- filter(sampleDataBacteria, sample %in% samples)
  
  alphaDivDFList[[i]] <- sampleDataFilt
  names(alphaDivDFList)[i] <- comparison
}
```

#### **Community Composition Plots** 

```{r comm-comp-biobank}
#----- UI -----#

inputPanel(
  # select phyloseq object
  selectInput(inputId = "commPSObjBB", label = "Comparison:",
              choices = c("All CRPS vs All BioBank" = "CRPS_BB",
                          "Chronic CRPS vs Chronic BioBank" = "Chronic_BB",
                          "Acute CRPS vs Acute BioBank" = "Acute_BB"),
              selected = "CRPS_BB"),
  # select taxonomic rank
  selectInput(inputId = "commTaxRankBB", label = "Taxonomic Rank",
              choices = c("Kingdom", "Phylum", "Class", "Order", "Family",
                          "Genus", "Species"),
              selected = "Phylum"),
  # select abundance filter
  sliderInput(inputId = "commAbdFilterBB", label = "Abundance Filter",
              min = 0, max = 1, step = 0.05, value = 0),
  # hide NA?
  checkboxInput(inputId = "commHideNABB", label = "Hide unnamed taxa", 
                value = FALSE)
)

# update plot:
actionButton(inputId = "commUpdateBB", label = "Update Composition Plot",
             icon = icon("fas fa-sync-alt"))

#----- React -----#

CalculateAbundanceBB <- eventReactive(input$commUpdateBB, {
  
  withProgress(message = "Calculating Abundance...", value = 1, {
                 abdDF <- MakeAbundanceDF(physeq = physeqSubsetList[[input$commPSObjBB]],
                                          taxRank = input$commTaxRankBB,
                                          abundanceFilter = input$commAbdFilterBB,
                                          pruneMissing = input$commHideNABB)

                 if (dim(abdDF)[1] == 0) {
                   stop("No taxa meet this filtering criteria. 
                        Try lowering the Abundance Filter option.",
                        call. = FALSE)
                 } else {
                   if (input$commTaxRankBB == "Species") {
                     
                     abdDF <- mutate(abdDF,
                                     "GenusSpecies" = paste(Genus, Species, 
                                                            sep = " "),
                                     no = "other")
                   }
                 }
                 abdDF
               })
}, ignoreNULL = FALSE)

GetFacetString <- eventReactive(input$commUpdateBB, {
  
  facetString <- ifelse(input$commPSObjBB == "CRPS_BB",
                        yes = "~ sample_type",
                        no = "~ crps_match")
  
  return(facetString)
  
}, ignoreNULL = FALSE)

renderPlotly({
  compPlot <- PlotCommunityComposition(abdDF = CalculateAbundanceBB(),
                                       taxRank = ifelse(isolate(input$commTaxRankBB) == "Species",
                                                        yes = "GenusSpecies",
                                                        no = isolate(input$commTaxRankBB)),
                                       facetFormula = GetFacetString())
  
  ggplotly(compPlot, height = 600, width = 1000)
})

```

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>


#### **Alpha Diversity**

```{r alpha-div-biobank}
#----- UI -----#

inputPanel(
  selectInput(inputId = "alphaDivBB", label = "Comparison:",
              choices = c("All CRPS vs All BioBank" = "CRPS_BB",
                          "Chronic CRPS vs Chronic BioBank" = "Chronic_BB",
                          "Acute CRPS vs Acute BioBank" = "Acute_BB"),
              selected = "CRPS_BB")
)

#----- React -----#

GetColVarAlpha <- reactive({
  colVar <- ifelse(input$alphaDivBB == "CRPS_BB",
                   yes = "sample_type",
                   no = "crps_match")
  return(colVar)
})

alphaDivPlotsBB <- reactive({
  pmap(.l = list(yVar = alphaIdxList,
                 yLab = names(alphaIdxList)),
       .f = PlotAlphaDiversity,
       df = alphaDivDFList[[input$alphaDivBB]],
       xVar = isolate(GetColVarAlpha()))
})

renderPlot({
  grid.arrange(alphaDivPlotsBB()[["Richness"]],
               alphaDivPlotsBB()[["Shannon Diversity"]],
               alphaDivPlotsBB()[["Faith's PD"]],
               ncol = 2, nrow = 2)
}, width = 900, height = 900)
```

#### **Beta Diversity**

```{r beta-div-BB}
#----- UI -----#

inputPanel(
  # select physeq object
  selectInput(inputId = "betaPhyseqBB", label = "Comparison:",
              choices = c("All CRPS vs All BioBank" = "CRPS_BB",
                          "Chronic CRPS vs Chronic BioBank" = "Chronic_BB",
                          "Acute CRPS vs Acute BioBank" = "Acute_BB"),
              selected = "CRPS_BB"),
  # select distance method
  radioButtons(inputId = "betaDistBB", label = "Distance Method:",
               choices = c("UniFrac" = "unifrac",
                           "Weighted UniFrac" = "wunifrac"),
               selected = "unifrac"),
  # show sample names
  checkboxInput(inputId = "betaShowNamesBB", label = "Show point labels",
                value = FALSE),
  # show confidence ellipses
  checkboxInput(inputId = "betaShowEllipsesBB", 
                label = "Show confidence ellipses", value = TRUE)
)

#----- React -----#

GetColVarBeta <- reactive({
  colVar <- ifelse(input$betaPhyseqBB == "CRPS_BB",
                   yes = "sample_type",
                   no = "crps_match")
})

CalculateOrdinationBB <- reactive({
  withProgress(message = "Calculating ordination (BioBank controls)...", 
               value = 1, {
    # Generate ordination object
    set.seed(28306097)
    ordinationObj <- ordinate(physeqSubsetList[[input$betaPhyseqBB]], 
                              method = "PCoA",
                              distance = input$betaDistBB)
  })
})

renderPlot({
  basePlot <- plot_ordination(physeqSubsetList[[isolate(input$betaPhyseqBB)]],
                              ordination = CalculateOrdinationBB(),
                              color = isolate(GetColVarBeta()), 
                              shape = isolate(GetColVarBeta())) +
    theme_bw() +
    theme(legend.title = element_text(size = 18),
          legend.text = element_text(size = 16),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          axis.text.x = element_text(size = 12),
          axis.text.y = element_text(size = 12)) +
    geom_point(size = 4)
  
  if (input$betaShowNamesBB) {
    basePlot <- basePlot +
      geom_text_repel(aes_string(label = "sample"), color = "grey30", size = 4)
  }
  
  if (input$betaShowEllipsesBB == TRUE) {
    basePlot <- basePlot + stat_ellipse(type = "norm")
  }
  
  basePlot
  
}, height = 600, width = 900)

```

**ADONIS Results for selected Comparison and Distance Measure:**

```{r beta-adonis-BB}
set.seed(28306097)
renderPrint(
  RunAdonis(physeqObj = physeqSubsetList[[isolate(input$betaPhyseqBB)]],
            category = GetColVarBeta(), 
            distance = input$betaDistBB)
)
```

<br>

### Session Info

```{r session-info}
Sys.Date()
getwd()
sessionInfo()
```
  