---
title: "Human CRPS Combined HHC Analyses"
author: "Rachel Rodgers"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=8,
                      fig.height=6,
                      warning = FALSE,
                      message = FALSE)
```

```{r load-libraries, message=FALSE, warning=FALSE}
library("phyloseq")
library("readxl")
library("microbiome")
library("vegan")
library("ggpubr")
library("ggrepel")
library("DESeq2")
library("data.table")
library("scales")
library("tidyverse")

source("shared_R_scripts/Helper_Functions.R")
source("shared_R_scripts/BoxPlotHelpers.R")
source("shared_R_scripts/BiomarkerHelpers.R")
```

```{r function-definitions}
RunAdonis <- function(physeqObj, category, distance) {
  bdist <- phyloseq::distance(physeqObj, distance)
  col <- as(sample_data(physeqObj), "data.frame")[, category]
  # Adonis test
  adonis.bdist <- adonis(bdist ~ col)
  return(adonis.bdist)
}
```

## Overview {.tabset}

In this HHC analysis, we want to determine if HHCs with significant daily pain ("Pain") and HHCs without significant daily pain ("No Pain") have different stool bacterial microbiomes.

```{r read-in-data, results='hide'}
# physeqBacteria comes from rarefied data, see script 
#   201129_Human_CRPS_Combined_16S_Analysis.Rmd
physeqBacteria <- readRDS("../data/physeqObjects/physeqBacteria.RDS")
physeqBacteria # 1593 x 64

# read-in HHC pain data
hhcPain <- read_excel("../documents/original_documents/HHC_pain.xlsx") %>% 
  select(sample, Sample_Status) %>% 
  dplyr::rename(HHC_Sample_Status = Sample_Status)

hhcSubgroupSamples <- hhcPain$sample

# add this information to the physeqBacteria sample data
sampleData <- as(sample_data(physeqBacteria), "data.frame")

# Edit the sample data to identify households with HHC that report pain ("bad" households)
sampleDataModified <- merge(hhcPain, sampleData, by = "sample", all = TRUE) 

badHouseholds <- sampleDataModified %>% 
  filter(sample %in% hhcSubgroupSamples,
         HHC_Sample_Status == "Pain") %>% 
  pull(household)

# add "bad_household" boolean column, and for non-biobank controls, a designation
#   for HHC good/bad household and CRPS good/bad household
sampleDataModified <- sampleDataModified %>% 
  mutate("bad_household" = ifelse(household %in% badHouseholds,
                                  yes = TRUE, no = FALSE),
         "status_household_description" = case_when(sample_status == "healthy" &
                                                      !bad_household ~ "HHC_good_household",
                                                    sample_status == "healthy" &
                                                      bad_household ~ "HHC_bad_household",
                                                    sample_status == "CRPS" &
                                                      !bad_household ~ "CRPS_good_household",
                                                    sample_status == "CRPS" &
                                                      bad_household ~ "CRPS_bad_household")) %>% 
  select("bad_household", "status_household_description", everything())

row.names(sampleDataModified) <- sampleDataModified$sample

# 221122 - Remove sample Baldridge007B as per Lara's request
sampleDataModified <- sampleDataModified %>% 
  filter(sample != "Baldridge007B")

sample_data(physeqBacteria) <- sample_data(sampleDataModified)
```

```{r save-physeqBacteria-with-modified-sample-data, eval=FALSE, include=FALSE}
saveRDS(physeqBacteria, "../data/physeqObjects/physeqBacteriaModifiedMetadata.RDS")
```

```{r write-out-lefse-input-data, eval=FALSE, include=FALSE}
# as per Lara's request Nov 29, 2022
source("./shared_R_scripts/GenerateLefseData.R")

# CRPS bad households vs CRPS good households
crpsSamples <- sampleDataModified %>% 
  filter(sample_type == "CRPS") %>% 
  pull(sample)
physeqCRPS <- prune_samples(samples = crpsSamples, x = physeqBacteria) %>% 
  RemoveMissingTaxa() # 968 x 22

crpsLefseIn <- GenerateLefseData(physeqObject = physeqCRPS,
                                 categoryColumnName = "status_household_description",
                                 sampleColumnName = "sample")
write.table(x = crpsLefseIn, file = "../data/LEfSe_input/crps_goodHH_badHH.txt",
            quote = FALSE, sep = "\t", row.names = TRUE, col.names = FALSE)

# HHC bad households vs HHC good households
hhcSamples <- sampleDataModified %>% 
  filter(sample_type == "HHC") %>% 
  pull(sample)
physeqHHC <- prune_samples(samples = hhcSamples, x = physeqBacteria) %>% 
  RemoveMissingTaxa() # 961 x 22

hhcLefseIn <- GenerateLefseData(physeqObject = physeqHHC,
                                categoryColumnName = "status_household_description",
                                sampleColumnName = "sample")
write.table(x = hhcLefseIn, file = "../data/LEfSe_input/hhc_goodHH_badHH.txt",
            quote = FALSE, sep = "\t", row.names = TRUE, col.names = FALSE)
```

```{r physeqSubsetList, results='hide'}
#----- HHC Samples Only -----#
physeqHHC <- physeqBacteria %>% 
  subset_samples(sample %in% hhcSubgroupSamples) %>% 
  RemoveMissingTaxa()
physeqHHC # 961 x 22; group by HHC_Sample_Status

#----- "Good Households" -----#
# subset the physeq object to keep both HHC and CRPS samples for "good" and "bad"
#   households only (no biobank controls)
physeqGood <- physeqBacteria %>% 
  subset_samples(sample_type != "BioBank_Control" & !bad_household) %>% 
  RemoveMissingTaxa()
physeqGood # 1115 x 24

#----- "Bad Households" -----#
physeqBad <- physeqBacteria %>% 
  subset_samples(sample_type != "BioBank_Control" & bad_household) %>% 
  RemoveMissingTaxa()
physeqBad # 729 x 20

#----- No BioBank Controls -----#
physeqNoBB <- physeqBacteria %>% 
  subset_samples(sample_type != "BioBank_Control") %>% 
  RemoveMissingTaxa()
physeqNoBB # 1383 x 44

#----- BB vs HHC -----#
physeqControls <- physeqBacteria %>% 
  subset_samples(sample_type %in% c("BioBank_Control", "HHC")) %>% 
  RemoveMissingTaxa()
physeqControls # 1235 x 41

#----- BB vs all CRPS -----#
physeqCRPSBB <- physeqBacteria %>% 
  subset_samples(sample_type %in% c("BioBank_Control", "CRPS")) %>% 
  RemoveMissingTaxa()
physeqCRPSBB # 1236 x 41

#----- CRPS Bad HH vs CRPS Good HH -----#
physeqCRPS <- physeqBacteria %>% 
  subset_samples(sample_type == "CRPS") %>% 
  RemoveMissingTaxa()
physeqCRPS # 968 x 22

physeqSubsetList <- list("HHC" = physeqHHC, 
                         "Good_Households" = physeqGood,
                         "Bad_Households" = physeqBad,
                         "Controls" = physeqControls,
                         "No_BB" = physeqNoBB,
                         "BB_vs_CRPS" = physeqCRPSBB,
                         "CRPS" = physeqCRPS)

# variables to split each group in physeqSubsetList into:
groupVars <- c("HHC" = "HHC_Sample_Status",
               "Good_Households" = "sample_status",
               "Bad_Households" = "sample_status",
               "Controls" = "sample_type",
               "No_BB" = "status_household_description",
               "BB_vs_CRPS" = "sample_type",
               "CRPS" = "status_household_description")
```

### Species-level Community Composition

```{r community-composition}
# Calculate community composition for all samples at the species level
speciesAbdDF <- MakeAbundanceDF(physeq = physeqBacteria,
                                taxRank = "Species",
                                abundanceFilter = 0)

# Edit species column to be genus species
speciesAbdDFModified <- speciesAbdDF %>% 
  mutate(species = ifelse(is.na(Species), 
                          yes = paste(Genus, "spp."),
                          no = paste(Genus, Species)),
         species = ifelse(species == "NA spp.",
                          yes = "unclassified",
                          no = species)) %>% 
  select(species, Abundance, everything())

# Filter out things with an abundance < 10%
speciesAbdFiltered <- speciesAbdDFModified %>% 
  filter(Abundance > 0.15)

# Standardize colors between plots
uniqueSpecies <- unique(speciesAbdFiltered$species)
speciesColors <- hue_pal()(length(uniqueSpecies))
names(speciesColors) <- uniqueSpecies

# Community plot by sample_type
communityBySampleType <- PlotCommunityComposition(abdDF = speciesAbdFiltered,
                                                  taxRank = "species",
                                                  facetFormula = "~ sample_type",
                                                  facetCol = 2, facetRow = 2)
communityBySampleType <- communityBySampleType +
  scale_fill_manual(values = speciesColors) +
  theme_bw() + 
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank())
communityBySampleType

# Community plot by status_household_description
communityByHH <- PlotCommunityComposition(abdDF = speciesAbdFiltered,
                                          taxRank = "species",
                                          facetFormula = "~ status_household_description",
                                          facetCol = 2, facetRow = 2)
communityByHH <- communityByHH + 
  scale_fill_manual(values = speciesColors) +
  theme_bw() + 
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank())
communityByHH
```

```{r save-community-plots, eval=FALSE}
figPath <- "../analysis/figures"
dir.create(file.path(figPath, "community_composition"))

communityPlots <- list("communityPlotBySampleType" = communityBySampleType,
                       "communityPlotByHH" = communityByHH)

figPaths <- map(.x = names(communityPlots),
                .f = ~ paste0(file.path(figPath, "community_composition"),
                              "/", .x, "_filt15Perc.jpg"))

names(figPaths) <- names(communityPlots)

walk2(.x = figPaths, .y = communityPlots,
      .f = ~ ggsave(filename = .x, plot = .y, width = 11, height = 8.5,
                    units = "in"))
```

```{r phylum-community-composition}
phylumAbdDFList <- map(.x = physeqSubsetList,
                       .f = ~ MakeAbundanceDF(physeq = .x,
                                              taxRank = "Phylum",
                                              abundanceFilter = 0))

facetFormulas <- paste("~", groupVars)

phylumCommunityPlots <- pmap(.l = list(abdDF = phylumAbdDFList,
                                       facetFormula = facetFormulas),
                             .f = PlotCommunityComposition,
                             facetRow = 2, facetCol = 2)

communityPlotsAnno <- map2(.x = phylumCommunityPlots, 
                           .y = names(phylumCommunityPlots),
                           .f = ~ annotate_figure(p = .x, top = .y))

communityPlotsAnno
```

### Bacteroides:Firmicutes Ratio

```{r bf-ratio}
selectPhylaAbdDFList <- map(.x = phylumAbdDFList,
                            .f = ~ .x %>% 
                              filter(Phylum %in% c("Bacteroidetes",
                                                   "Firmicutes")) %>% 
                              select(Phylum, Abundance, sample, 
                                     HHC_Sample_Status, sample_status,
                                     status_household_description,
                                     sample_type))
  
phylaAbdPivotList <- map(.x = selectPhylaAbdDFList,
                         .f = ~ .x %>% 
                           pivot_wider(names_from = Phylum,
                                       values_from = Abundance) %>% 
                           mutate(ratio = Bacteroidetes/Firmicutes))

ratioPlotList <- pmap(.l = list(df = phylaAbdPivotList,
                                xVar = groupVars,
                                statMethod = c(rep("wilcox.test", 4), 
                                               "kruskal.test", 
                                               rep("wilcox.test", 2))),
                      .f = MakeBoxPlot,
                      yVar = "ratio",
                      label_y = "Bacteroidtes:Firmicutes (rel. abd.)")

ratioPlotListAnno <- map2(.x = ratioPlotList,
                          .y = names(ratioPlotList),
                          .f = ~ annotate_figure(p = .x, top = .y))

ratioPlotListAnno
```

```{r save-b-f-ratio-for-lara, eval=FALSE}
fileList <- paste0("../analysis/", names(phylaAbdPivotList),
                   "_bacteroidetes_firmicutes_ratio.txt")
names(fileList) <- names(phylaAbdPivotList)
walk2(.x = phylaAbdPivotList, .y = fileList,
      .f = write.table,
      quote = FALSE, sep = "\t", row.names = FALSE)
```

### Alpha Diversity

```{r calculate-alpha-idx, results='hide'}
CalculateAlphaDiversity <- function(physeqObj) {
  sampleData <- as(sample_data(physeqObj), "data.frame")
  alphaDiv <- microbiome::alpha(physeqObj, index = c("observed",
                                                     "diversity_shannon"))
  sampleDataWithAlpha <- merge(alphaDiv, sampleData, by = "row.names", all = TRUE)
}

alphaDivList <- map(.x = physeqSubsetList, .f = ~ CalculateAlphaDiversity(.))

richnessPlots <- pmap(.l = list(df = alphaDivList,
                                xVar = groupVars,
                                alphaPlotTitle = paste(names(alphaDivList), 
                                                       "Richness"),
                                statMethod = c(rep("wilcox.test", 4),
                                               "kruskal.test",
                                               rep("wilcox.test", 2))),
                      .f = PlotAlphaDiversity,
                      yVar = "observed",
                      yLabel = "Richness")

diversityPlots <- pmap(.l = list(df = alphaDivList,
                                xVar = groupVars,
                                alphaPlotTitle = paste(names(alphaDivList), 
                                                       "Shannon Diversity"),
                                statMethod = c(rep("wilcox.test", 4),
                                               "kruskal.test",
                                               rep("wilcox.test", 2))),
                      .f = PlotAlphaDiversity,
                      yVar = "diversity_shannon",
                      yLabel = "Diversity")

alphaPlotsArr <- map2(.x = richnessPlots, .y = diversityPlots,
                      .f = ~ ggarrange(.x, .y, ncol = 2, nrow = 1))

alphaPlotsArr
```

```{r save-alphaDivList, eval=FALSE, include=FALSE}
write.table(x = alphaDivList$No_BB,
            file = "../analysis/alpha_diversity_values.txt",
            quote = FALSE, sep = "\t", row.names = FALSE)

write.table(x = alphaDivList$Controls,
            file = "../analysis/alpha_diversity_values_controls.txt",
            quote = FALSE, sep = "\t", row.names = FALSE)
```

### Beta Diversity

```{r beta-diversity, warning=FALSE}
#----- calculate unweighted & weighted ordinations -----#
set.seed(4217589)
unweightedOrdList <- map(.x = physeqSubsetList,
                         .f = ordinate,
                         method = "PCoA", distance = "unifrac")
set.seed(4217589)
weightedOrdList <- map(.x = physeqSubsetList,
                       .f = ordinate,
                       method = "PCoA", distance = "wunifrac")

#----- Run ADONIS & Extract P-Value -----#
set.seed(4217589)
unweightedADONISList <- map2(.x = physeqSubsetList,
                             .y = groupVars,
                             .f = RunAdonis,
                             distance = "unifrac")
unweightedPVals <- map(.x = unweightedADONISList,
                       .f = ~ .x$aov.tab$`Pr(>F)`[1])

set.seed(4217589)
weightedADONISList <- map2(.x = physeqSubsetList,
                           .y = groupVars,
                           .f = RunAdonis,
                           distance = "wunifrac")
weightedPVals <- map(.x = weightedADONISList,
                     .f = ~ .x$aov.tab$`Pr(>F)`[1])

#----- Generate Ordination Plots -----#
unweightedOrdPlots <- pmap(.l = list(ordObj = unweightedOrdList,
                                     physeqObj = physeqSubsetList,
                                     mySubtitle = paste("ADONIS p:",
                                                        unweightedPVals),
                                     colorVar = groupVars),
                           .f = MakeOrdinationPlot,
                           colorValues = cbPaletteGrey,
                           myTitle = "Unweighted UniFrac")

weightedOrdPlots <- pmap(.l = list(ordObj = weightedOrdList,
                                   physeqObj = physeqSubsetList,
                                   mySubtitle = paste("ADONIS p:", 
                                                      weightedPVals),
                                   colorVar = groupVars),
                         .f = MakeOrdinationPlot,
                         colorValues = cbPaletteGrey,
                         myTitle = "Weighted UniFrac")

#----- Format Ordination Plots -----#
groups <- names(physeqSubsetList)

finalOrdPlots <- vector(mode = "list", length = length(groups))

for (i in 1:length(groups)) {
  currentGroup <- groups[i]
  # add ellipses to weighted & unweighted UniFrac plots
  plotList <- list("unweighted" = unweightedOrdPlots,
                   "weighted" = weightedOrdPlots)
  
  plotsWithEllipses <- map(.x = plotList,
                           .f = ~ .x[[currentGroup]] + 
                             stat_ellipse(type = "norm"))
  
  # arrange/annotate plots
  plotsArr <- ggarrange(plotsWithEllipses$unweighted,
                        plotsWithEllipses$weighted,
                        common.legend = TRUE,
                        nrow = 1, ncol = 2, legend = "bottom")
  plotsFinal <- annotate_figure(p = plotsArr,
                                top = currentGroup)
  
  # add to list
  finalOrdPlots[[i]] <- plotsFinal
  names(finalOrdPlots)[i] <- currentGroup
}
```

```{r ordPlotList}
finalOrdPlots
```

### Biomarker Analysis with DESeq2

Checking for biomarkers that distinguish between these pairs of groups:

  * BioBank vs CRPS
  * HHC vs CRPS 
  * Acute CRPS vs Chronic CRPS
  * Good CRPS vs Bad CRPS
  
```{r subset-physeq-for-deseq-analysis}
# BioBank versus CRPS already generated (physeqSubsetList$BB_vs_CRPS)

#----- HHC vs CRPS -----#
physeqHHCvsCRPS <- physeqBacteria %>% 
  subset_samples(sample_type %in% c("CRPS", "HHC")) %>% 
  RemoveMissingTaxa() # 1383 x 44

#----- Acute CRPS vs Chronic CRPS -----#
physeqCRPSType <- physeqBacteria %>% 
  subset_samples(description %in% c("Acute", "Chronic")) %>% 
  RemoveMissingTaxa() # 968 x 22

#----- CRPS Good HH vs CRPS Bad HH -----#
physeqCRPSHH <- physeqBacteria %>% 
  subset_samples(status_household_description %in% 
                   c("CRPS_bad_household", "CRPS_good_household")) %>% 
  RemoveMissingTaxa() # 968 x 22
  
#----- store these new physeq objects in list for easier processing -----#
deseqPhyseqList <- list("BB vs CRPS" = physeqSubsetList$BB_vs_CRPS,
                        "HHC vs CRPS" = physeqHHCvsCRPS,
                        "Acute vs Chronic CRPS" = physeqCRPSType,
                        "CRPS Bad HH vs CRPS Good HH" = physeqCRPSHH)
```
  
```{r run-deseq-analyses}
physeqNames <- names(deseqPhyseqList)

variableList <- list("sample_type", 
                     "sample_type", 
                     "description", 
                     "status_household_description")
names(variableList) <- physeqNames

numeratorList <- list("BioBank_Control",
                      "HHC",
                      "Acute",
                      "CRPS_bad_household")
names(numeratorList) <- physeqNames

denominatorList <- list("CRPS",
                        "CRPS",
                        "Chronic",
                        "CRPS_good_household")
names(denominatorList) <- physeqNames

#----- Modify GenerateDESeqResults fxn for work-around -----#

GenerateDESeqResultsModified <- function(physeq, variable, numerator, denominator) {
  
  # Returns DESeq Results as Formal Class "DESeqResults"
  # Create formula from string variable
  formula <- as.formula(paste("~", variable, sep = " "))
  # Convert to deseq data set object
  ds <- phyloseq_to_deseq2(physeq, design = formula)
  # work-around
  counts <- counts(ds)
  geoMeans <- apply(counts, 1,
                      function(row) {if (all(row == 0)) 0 else exp(mean(log(row[row != 0])))})
  dds <- estimateSizeFactors(ds, geoMeans = geoMeans)
  # Run analysis
  ddsAnalysis <- DESeq(dds, test = "Wald", fitType = "local", betaPrior = FALSE)
  # Extract and format results
  ddsResults <- results(ddsAnalysis,
                        contrast = c(variable, numerator, denominator)) 
}

#----- Run DESeq2 -----#
deseqResults <- pmap(.l = list(physeq = deseqPhyseqList,
                               variable = variableList,
                               numerator = numeratorList,
                               denominator = denominatorList),
                     .f = GenerateDESeqResultsModified)

deseqResultTables <- map2(.x = deseqPhyseqList,
                          .y = deseqResults,
                          .f = ~ GenerateDESeqResultsTable(physeq = .x,
                                                           ddsResults = .y),
                          sigThreshold = 0.05)
```

```{r generate-deseq-volcano-plots}
volcanoPlots <- pmap(.l = list(physeq = deseqPhyseqList,
                               resultsDataTable = deseqResultTables,
                               plotTitle = names(deseqPhyseqList)),
                     .f = PlotStaticVolcano,
                     sigThreshold = 0.05)

volcanoPlots
```

